---
title: "Spatial Analysis Project"
author: "Chung-Wing Ko"
date: "25 April 2022"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Background Information

My research project focuses on measuring belowground (soil and root) characteristics of the Central Catchment Nature Reserve in Singapore. I'm interested in how land use history has created these forest types, ranging from near-primary patches to secondary forests of various successional stages. The Central Catchment has not been extensively surveyed before for belowground traits, which are important to understand for future restoration efforts. Informed restoration should incorporate soil environment into tree planting and other initiatives, which may influence the success of these reforestation efforts. 

My project is in collaboration with NParks, who are reestablishing 62 plots previously established in 1992 by Wong and colleagues. NParks is focusing on aboveground traits such as tree composition and understory composition, while I focus on belowground traits. For this project, I hope to interpolate soil nitrogen concentration across the lower Central Catchment, and have currently sampled 18 of the 62 plots total. 

Soil nitrogen is an important nutrient for plant growth, and is hypothesised to be one of the limiting factors in plant success, alongside phosphorus. This variable is therefore of significant interest to see its relative abundance across the Central Catchment and observe any potential spatial patterns in its distribution. 

## Load Libraries and Data
```{r}
#load libraries
#data cleaning
library(tidyr)
library(dplyr)
library(stringr)
#plotting
library(ggplot2)
library(MetBrewer) #color palette
library(tmap)
#spatial analysis
library(sf)
library(sp)
library(spatstat)
library(maptools)
library(rgdal)
library(raster)
library(gstat) #IDW
library(geoR)

#load plot coordinates
coords <- read.csv("/Users/chungwingko/Desktop/CCProject/coordinates - Sheet1.csv")
#load boundary
boundary <- st_read("/Users/chungwingko/Desktop/CCProject/Map/CCcutout.shp")
```

## Data Cleaning

These data are various measurements collected from the 18 plots sampled thus far. Measurements of soil nitrogen must be adjusted for factors of extractant volume and soil moisture, so all data is combined and adjusted before soil nitrogen is isolated for this class project.

### Attribute Data
```{r}
#load + clean root mass data
rootbiomass <- read.csv("/Users/chungwingko/Desktop/CCProject/DryRootMasses - Sheet1.csv")
rootbiomass <- rootbiomass %>% dplyr::select(ID, FT, Necromass, Total.root.biomass)
rootbiomass$PercentNecro <- rootbiomass$Necromass/rootbiomass$Total.root.biomass
rootbiomass <- rootbiomass[-c(91:98),]
rootbiomass$FT <- as.factor(rootbiomass$FT)

#load + clean TOC/TN data
TOC <- read.csv("/Users/chungwingko/Desktop/CCProject/TOC data - Sheet1.csv")
#already subtracted all blanks
#corrected W05-F-B by using W16-F-B (run same week)
#corrected W57-F-B by using 56-F-B (same week)
TOC <- TOC %>% dplyr::select(Sample.ID, TOC_corrected, TC_corrected, IC_corrected, TN_corrected) %>% drop_na()
#assigning NA to outliers: W06-F-C (all), W07-F-C (TOC, TC, TN), W07-NF-SE (TOC, TC, TN), W57-F-SE (TOC, TC), W21-F-SW (TOC, TC, TN)
TOC[TOC$Sample.ID == "W06-F-C",2:5] <- NA
TOC[TOC$Sample.ID == "W07-F-C", c(2, 3, 5)] <- NA
TOC[TOC$Sample.ID == "W07-NF-SE", c(2, 3, 5)] <- NA
TOC[TOC$Sample.ID == "W57-F-SE", c(2:3)] <- NA
TOC[TOC$Sample.ID == "W21-F-SW", c(2, 3, 5)] <- NA

TOC_F <- TOC[str_detect(TOC$Sample.ID, "-F-"),]
TOC_F <- TOC_F %>% rename(TOC.F = TOC_corrected,
                          TC.F = TC_corrected,
                          IC.F = IC_corrected,
                          TN.F = TN_corrected)
TOC_F$Sample.ID <- gsub("-F", "", as.character(TOC_F$Sample.ID))
TOC_F <- TOC_F %>% filter(!str_detect(Sample.ID, "-B"))

TOC_NF <- TOC[str_detect(TOC$Sample.ID, "-NF-"),]
TOC_NF <- TOC_NF %>% rename(TOC.NF = TOC_corrected,
                          TC.NF = TC_corrected,
                          IC.NF = IC_corrected,
                          TN.NF = TN_corrected)
TOC_NF$Sample.ID <- gsub("-NF", "", as.character(TOC_NF$Sample.ID))
TOC_NF <- TOC_NF %>% filter(!str_detect(Sample.ID, "-B"))

TOC_TN <- left_join(TOC_F, TOC_NF, by = c("Sample.ID"))
TOC_TN[TOC_TN < 0] <-0 
TOC_TN <- TOC_TN %>% rename(ID = Sample.ID)

#import soil moisture data
soil_moisture <- read.csv("/Users/chungwingko/Desktop/CCProject/FreshSoilExtracts.xlsx - Inorg_microbial N_P_soil_mass.csv")
soil_moisture <- soil_moisture %>% rename(KCl.F = KCl.Fum.Soil.mass..g.,
                          KCl.NF = KCl.non.fum.Soil.mass..g.,
                          Bicarb.F = Bicarb.Fum.Soil.mass..g.,
                          Bicarb.NF = Bicarb.non.fum.Soil.mass..g.,
                          Percent.Dry.Soil = X..Dry.Soil)
soil_moisture <- soil_moisture %>% dplyr::select(ID, KCl.F, KCl.NF, Bicarb.F, Bicarb.NF, Percent.Dry.Soil)
soil_moisture$KCl.F <- soil_moisture$KCl.F * soil_moisture$Percent.Dry.Soil
soil_moisture$KCl.NF <- soil_moisture$KCl.NF * soil_moisture$Percent.Dry.Soil
soil_moisture$Bicarb.F <- soil_moisture$Bicarb.F * soil_moisture$Percent.Dry.Soil
soil_moisture$Bicarb.NF <- soil_moisture$Bicarb.NF * soil_moisture$Percent.Dry.Soil

#load + clean SEAL data
SEAL <- read.csv("/Users/chungwingko/Desktop/CCProject/Cleaned SEAL data - Sheet1.csv")
SEAL <- SEAL %>% dplyr::select(ID, NH3_corrected, NO3_corrected) #blanks subtracted
SEAL <- SEAL %>% rename(NH3 = NH3_corrected,
                        NO3 = NO3_corrected)

SEAL_F <- SEAL[str_detect(SEAL$ID, "-F-"),]
SEAL_F <- SEAL_F %>% rename(NH3.F = NH3,
                          NO3.F = NO3)
SEAL_F$ID <- gsub("-F", "", as.character(SEAL_F$ID))
SEAL_F <- SEAL_F %>% filter(!str_detect(ID, "-B"))

SEAL_NF <- SEAL[str_detect(SEAL$ID, "-NF-"),]
SEAL_NF <- SEAL_NF %>% rename(NH3.NF = NH3,
                          NO3.NF = NO3)
SEAL_NF$ID <- gsub("-NF", "", as.character(SEAL_NF$ID))
SEAL_NF <- SEAL_NF %>% filter(!str_detect(ID, "-B"))

SEAL <- left_join(SEAL_F, SEAL_NF, by = c("ID"))
SEAL[SEAL < 0] <-0 

#join all datasets
CCProject <- left_join(TOC_TN, rootbiomass, by = "ID")
CCProject <- left_join(CCProject, SEAL, by = "ID")
CCProject <- left_join(CCProject, soil_moisture, by = "ID")
#separate plot and direction
CCProject <- CCProject %>% separate(ID, c("Plot", "Direction"))
#change variable order
CCProject <- CCProject %>% dplyr::select(FT, Plot, Direction,Total.root.biomass, PercentNecro, TOC.F, TOC.NF, TC.F, TC.NF, IC.F, IC.NF, TN.F, TN.NF, NH3.F, NH3.NF, NO3.F, NO3.NF, KCl.F, KCl.NF, Bicarb.F, Bicarb.NF)

#adjust all values for dry soil mass 
CCProject$TOC.F = CCProject$TOC.F*25/CCProject$KCl.F
CCProject$TOC.NF = CCProject$TOC.NF*25/CCProject$KCl.NF
CCProject$TC.F = CCProject$TC.F*25/CCProject$KCl.F
CCProject$TC.NF = CCProject$TC.NF*25/CCProject$KCl.NF
CCProject$IC.F = CCProject$IC.F*25/CCProject$KCl.F
CCProject$IC.NF = CCProject$IC.NF*25/CCProject$KCl.NF
CCProject$TN.F = CCProject$TN.F*25/CCProject$KCl.F
CCProject$TN.NF = CCProject$TN.NF*25/CCProject$KCl.NF
CCProject$NH3.F = CCProject$NH3.F*25/CCProject$KCl.F
CCProject$NH3.NF = CCProject$NH3.NF*25/CCProject$KCl.NF
CCProject$NO3.F = CCProject$NO3.F*25/CCProject$KCl.F
CCProject$NO3.NF = CCProject$NO3.NF*25/CCProject$KCl.NF

CCProject$MicrobeC = (CCProject$TOC.F - CCProject$TOC.NF)/0.45
CCProject$microbeN = (CCProject$TN.F - CCProject$TN.NF)/0.54
```

### Filtering Soil Nitrogen Data

For this project, I'll be focusing on interpolating soil nitrogen content across the lower Central Catchment region. Therefore, I'll be using just the soil nitrogen measurements (in ppm), and all other data were used to help calculate concentrations and will not be needed. At each plot, 5 samples were taken and analysed separately. However, locations are only known on the plot level, so will be averaged across each plot and spatially analysed on that resolution. In the future, I will get more exact GPS locations of each sampling location from NParks, and will therefore be able to interpolate on a higher spatial resolution. 
```{r}
#for this project, focusing on soil nitrogen content
attributes <- CCProject %>% dplyr::select(Plot, Direction, TN.NF) %>% drop_na()
attributes <- attributes %>% group_by(Plot) %>% summarise(N = mean(TN.NF)) #averaging measurements
```

### Join Data
```{r}
#join coordinates and attributes
plots <- left_join(coords, attributes, by = "Plot")

#clean data
plots = dplyr::select(plots, -Plot) #dropping plot number
plots.sf <- st_as_sf(plots, coords = c("y", "x"), crs = 4326)
attributes(plots.sf)
```

## Data Visualisation

### Using ggplot2
```{r}
#plot boundary and points
ggplot() +
  geom_sf(data = boundary, fill = "#F9EED9") +
  geom_sf(data = plots.sf, aes(color = N)) + scale_color_gradientn(colors = met.brewer("Hokusai2")) + theme_classic()
```

### Using tmap
```{r}
tmap_options(check.and.fix = TRUE)
#convert sf objects to SpatialPointsDataFrame
boundary.sp <- as(boundary, "Spatial")
plots.sf <- st_as_sf(plots, coords = c("y", "x"), crs = 4326)
plots.sp <- as(plots.sf, "Spatial")

class(boundary.sp)
class(plots.sp)

#replace point boundary with that of polygon
plots.sp@bbox <- boundary.sp@bbox

#round to 2 decimals
plots.sp$N <- round(plots.sp$N, digit = 2)

#plot
tm_shape(boundary.sp) + tm_polygons() +
  tm_shape(plots.sp) +
  tm_dots(col="N", palette = "PuBu", title="Nitrogen concentration (ppm)", size=0.5) +
  tm_text("N", auto.placement =  TRUE, xmod=.5, size = .8) +
  tm_legend(legend.outside=TRUE)
tmap_options(check.and.fix = TRUE)
```

## Inverse Distance Weighted (IDW)

The IDW method uses nearby locations to compute values for unsampled locations, weighting measurements based on proximity based on a universal power coefficient which applys equally to all coefficients. 

### Interpolate
```{r}
#create an empty grid
grd <- as.data.frame(spsample(plots.sp, "regular", n=50000))
names(grd) <- c("x", "y")
coordinates(grd) <- c("x", "y")
gridded(grd) <- TRUE  #create SpatialPixel object
fullgrid(grd) <- TRUE  #create SpatialGrid object

#add projection info to the empty grid
proj4string(plots.sp) <- proj4string(plots.sp)
proj4string(grd) <- proj4string(plots.sp)

#interpolate the grid cells using a power value of 2
plots.idw <- gstat::idw(N ~ 1, plots.sp, newdata=grd, idp=2.0)

#convert to raster object then clip to boundary
r <- raster(plots.idw)
r.m <- mask(r, boundary.sp)
```

### Plot
```{r}
#plot
tmap_mode("view") #interactive plot
tm_shape(r.m) + 
  tm_raster(n=10, palette = "PuBu", auto.palette.mapping = FALSE,
            title="Predicted nitrogen concentration (ppm)") + 
  tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

### Fine Tune Power Parameter
```{r}
#leave-one-out validation routine
IDW.out <- vector(length = length(plots.sp))
for (i in 1:length(plots.sp)) {
  IDW.out[i] <- gstat::idw(N ~ 1, plots.sp[-i,], plots.sp[i,], idp=2)$var1.pred
}

#plot the differences
OP <- par(pty="s", mar=c(4,3,0,0))
plot(IDW.out ~ plots.sp$N, asp=1, xlab="Observed", ylab="Predicted", pch=16,
       col=rgb(0,0,0,0.5))
abline(lm(IDW.out ~ plots.sp$N), col="red", lw=2,lty=2)
  abline(0,1)
par(OP)

#compute RMSE
sqrt( sum((IDW.out - plots.sp$N)^2) / length(plots.sp))

#change power to 1
IDW.out1 <- vector(length = length(plots.sp))
for (i in 1:length(plots.sp)) {
  IDW.out1[i] <- gstat::idw(N ~ 1, plots.sp[-i,], plots.sp[i,], idp=1)$var1.pred
}

#plot the differences
OP1 <- par(pty="s", mar=c(4,3,0,0))
plot(IDW.out1 ~ plots.sp$N, asp=1, xlab="Observed", ylab="Predicted", pch=16,
       col=rgb(0,0,0,0.5))
abline(lm(IDW.out1 ~ plots.sp$N), col="red", lw=2,lty=2)
  abline(0,1)
par(OP1)

#compute RMSE
sqrt( sum((IDW.out1 - plots.sp$N)^2) / length(plots.sp))

#change power to 0
IDW.out2 <- vector(length = length(plots.sp))
for (i in 1:length(plots.sp)) {
  IDW.out2[i] <- gstat::idw(N ~ 1, plots.sp[-i,], plots.sp[i,], idp=0)$var1.pred
}

#plot the differences
OP2 <- par(pty="s", mar=c(4,3,0,0))
plot(IDW.out2 ~ plots.sp$N, asp=1, xlab="Observed", ylab="Predicted", pch=16,
       col=rgb(0,0,0,0.5))
abline(lm(IDW.out2 ~ plots.sp$N), col="red", lw=2,lty=2)
  abline(0,1)
par(OP2)

#compute RMSE
sqrt( sum((IDW.out2 - plots.sp$N)^2) / length(plots.sp))
```

The error got smaller as the power coefficient decreased, which gives all points within the search radius equal weight such that unsampled locations represent just the mean value of all sampled points within the search radius. This could be due to a wide variety of reasons, including that the spatial heterogeneity of the site is high, so even plots close together can have large differences in soil nitrogen content. Additionally, the points are not evenly spaced, and could influence some of these patterns. Overall, this implies that distance does not have a significant role in influencing measured soil nitrogen patterns. 

### Cross Validation

We can generate a 95% confidence interval map to look at the interpolation model more visually.
```{r}
#create interpolated surface
img <- gstat::idw(N~1, plots.sp, newdata=grd, idp=3)
n   <- length(plots.sp)
Zi  <- matrix(nrow = length(img$var1.pred), ncol = n)

#remove a point then interpolate (n times for each point)
st <- stack()
for (i in 1:n){
  Z1 <- gstat::idw(N~1, plots.sp[-i,], newdata=grd, idp=0)
  st <- addLayer(st,raster(Z1,layer=1))
  # calculated pseudo-value Z at j
  Zi[,i] <- n * img$var1.pred - (n-1) * Z1$var1.pred
}

#jackknife estimator of parameter Z at location j
Zj <- as.matrix(apply(Zi, 1, sum, na.rm=T) / n )

#compute (Zi* - Zj)^2
c1 <- apply(Zi,2,'-',Zj)            #compute the difference
c1 <- apply(c1^2, 1, sum, na.rm=T ) #sum the square of the difference

#compute the confidence interval
CI <- sqrt(1/(n*(n-1)) * c1)

#create (CI / interpolated value) raster
img.sig <- img
img.sig$v <- CI /img$var1.pred 

#clip the confidence raster to polygon
r <- raster(img.sig, layer="v")
r.m <- mask(r, boundary.sp)

#plot the map
tm_shape(r.m) + tm_raster(n=7, palette = "PuBu", title="95% confidence interval (N content in ppm)") +
  tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

The confidence intervals can be interpreted as that amount higher and lower than the interpolation. Based on the IDW method, the higher uncertainties exist in the northern and western areas of the spatial extent. The lack of sampled sites in the western portion largely add to this uncertainty. 

## Polynomial Fits

### 1st Order
```{r}
#define polynomial equation
f.1 <- as.formula(N ~ x + y) 
 
#add X and Y to point data
plots.sp$x <- coordinates(plots.sp)[,1]
plots.sp$y <- coordinates(plots.sp)[,2]

#run the regression model
lm.1 <- lm(f.1, data=plots.sp)

#use the regression model output to interpolate the surface
dat.1st <- SpatialGridDataFrame(grd, data.frame(var1.pred = predict(lm.1, newdata=grd))) 

#clip the interpolated raster to polygon
r <- raster(dat.1st)
r.m <- mask(r, boundary.sp)

#plot the map
tm_shape(r.m) + 
  tm_raster(n=10, palette="PuBu", auto.palette.mapping=FALSE, 
            title="Predicted nitrogen concentration (ppm)") +
  tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

The first order polynomial shows an east-west trend in the soil nitrogen concentration, which may not necessarily be uniform (thus leading to the need for a second order analysis). 

### 2nd Order
```{r}
#define polynomial equation
f.2 <- as.formula(N ~ x + y + I(x*x)+I(y*y) + I(x*y))
 
#add X and Y to point data
plots.sp$x <- coordinates(plots.sp)[,1]
plots.sp$y <- coordinates(plots.sp)[,2]

#run the regression model
lm.2 <- lm(f.2, data=plots.sp)

#use the regression model output to interpolate the surface
dat.2nd <- SpatialGridDataFrame(grd, data.frame(var1.pred = predict(lm.2, newdata=grd))) 

#clip the interpolated raster to polygon
r <- raster(dat.2nd)
r.m <- mask(r, boundary.sp)

#plot the map
tm_shape(r.m) + 
  tm_raster(n=10, palette="PuBu", auto.palette.mapping=FALSE, 
            title="Predicted nitrogen concentration (ppm)") +
  tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

The second order polynomial shows a curvature pattern previously not shown in the first order polynomial, which seems fairly significant. 

## Kriging

### Create Variogram Model
```{r}
#define polynomial equation
f.1 <- as.formula(N ~ x + y) 

#experimental semivariogram cloud plot
var.smpl <- variogram(f.1, plots.sp, cloud = TRUE)
plot(var.smpl, fit.ranges = TRUE, fit.sills = TRUE)

#sample variogram (cutoff and width based on cloud plot)
var.smpl <- variogram(f.1, plots.sp, cloud = FALSE, cutoff=2.5, width=.1)

#calculate parameters
max(var.smpl$dist)/3 #range
head(var.smpl$gamma, n = 3)
(351.61825 + 314.28503 + 44.60731)/3 #nugget
tail(var.smpl$gamma, n = 5)
(4.843825 + 139.391291 + 78.058451  + 6.959237+ 277.066504)/5 #partial sill

#compute variogram model
dat.fit  <- fit.variogram(var.smpl, fit.ranges = FALSE, fit.sills = FALSE,
                          vgm(psill=101.2639, model="Exp", range=1, nugget=26.8369))
#assess fit
plot(var.smpl, dat.fit, xlim=c(0,2.5))
```

The cloud plot did not show a clear pattern, and there are some outliers in the variogram model, meaning the mathematical model may not be the most accurate. The default calculations for partial sill, nugget, and range are shown below. The final values used were adjusted after visual inspection, and the Exponential model was chosen to represent the trend. 

### Generate Kriged Surface

This variogram model is used to provide localised weighting parameters, compared with the IDW method which uses a universal power parameter. The final kriged surface can be shown below. Similar to the polynomial fits, this also shows an east-west trend, but introduces some dynamic shifts based on these localised weights. 
```{r}
#define model
f.1 <- as.formula(N ~ x + y) 

#perform interpolation
dat.krg <- krige(f.1, plots.sp, grd, dat.fit)

#convert kriged surface to a raster object for clipping
r <- raster(dat.krg)
r.m <- mask(r, boundary.sp)

#plot
tm_shape(r.m) + 
  tm_raster(n=10, palette="PuBu", auto.palette.mapping=FALSE, 
            title="Predicted nitrogen  concentration (ppm)") +
  tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

### Variance Map

This variance map allows us to visualise uncertainty in the interpolation. The highest variance can be seen in the southwestern portion of the map, which makes sense as there is a lack of sampled sites in that region. Lower variances can be observed where the plots are most tightly clustered. 
```{r}
r <- raster(dat.krg, layer="var1.var")
r.m <- mask(r, boundary.sp)

#plot
tm_shape(r.m) + 
  tm_raster(n=7, palette ="Reds",
            title="Variance map \n(in squared ppm)") + tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```

### 95% Confidence Interval Map

Similarly to the variance map, the 95% CI map allows us to visualise uncertainty (interpreted as the amount above and below). The higher uncertainties are also consistently in the southwestern region of the map, for the same reasons. Interestingly, the range in confidence intervals for the Kriging method appears to be larger than when compared to the IDW method, but without understanding the maths behind the Kriging method (which is not covered here) this cannot fully be explained. 
```{r}
r <- sqrt(raster(dat.krg, layer="var1.var")) * 1.96
r.m <- mask(r, boundary.sp)

tm_shape(r.m) + 
  tm_raster(n=7, palette ="Reds",
            title="95% CI map \n(in ppm)") +tm_shape(plots.sp) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```
